Ensemble Stability of Biological Networks

### 2014-09-12 paper.tex
reorganize

%------------------------------------------------------------------------------
% Beginning of journal.tex
%------------------------------------------------------------------------------
%
% AMS-LaTeX version 2 sample file for journals, based on amsart.cls.
%
%        ***     DO NOT USE THIS FILE AS A STARTER.      ***
%        ***  USE THE JOURNAL-SPECIFIC *.TEMPLATE FILE.  ***
%
% Replace amsart by the documentclass for the target journal, e.g., tran-l.
%
\documentclass{amsart}

@@

\section{Stability analysis of biological networks}

Consider an n-dimensional vector
$$
x(t) = (x_1(t), \ldots x_n(t)) = \vec{x}(t)
$$
whose components are solutions to the arbitrary first order system
$$
\frac{dx_i(t)}{dt} = F_i(\vec{x}(t)), \; (i=1,\ldots,n)
$$
where $F_i$ represent, potentially nonlinear, functions of the given vector of state variables. We can associate a directed graph $G_X$ that describes the manner in which each of the variables depends upon one another given by an adjacency matrix $\adj(G_X)$ where
 \begin{displaymath}
   \adj(G_X)_{ij} = \left\{
     \begin{array}{ll}
       1, & F_i \hbox{ depends on } x_j\\
       0, & F_i \hbox{ does not depend on } x_j
     \end{array}
   \right.
\end{displaymath}
For example, given the system $X = \{\{x_1,x_2\},\{F_1,F_2\}\}$
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}
we have the graph
\begin{center}
\begin{tikzpicture}[
every state/.style={draw=red!50,very thick,fill=red!20}]
\begin{dot2tex}[styleonly,codeonly,neato,mathmode]
digraph G {
d2ttikzedgelabels = true;
node [style="state"];
edge [lblstyle="auto",topath="bend left",style="line width=1.5pt"];
x_1 -> x_2;
x_2 -> x_1;
x_2 -> x_2 [topath="loop above"];
}
\end{dot2tex}
\end{tikzpicture}
\end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$

@@

\section{System graph and decomposition into strongly connected components}
We will show how the stability of the dynamical system associated to Jacobian $A$ is characterized by properties of the associated graph $G_A$. The characteristic polynomial $\chi(A)$ associated to $A$ can be factored according to the strongly connected components of the graph $G$. $${\chi(M)'(z) \over \chi(M)(z)} = \tr (zI - M)^{-1}$$ where, $\chi(M)(z) = \det (zI - M)$ is the chararacteristic polynomial of the matrix $M$ with $z$ a complex variable.  This formula is readily verified when $M$ is diagonal and has the nice property of being additive over decompositions of the matrix. Making the change of variables $x=\frac{1}{z}$, we have $$\tr (I - xM)^n = \sum_{n=0}^\infty x^n \sum_{a \in \Vertex(G)} \sum_{p \in \Path_n (G)(a,a)} F(p)$$
where $G$ is the graph of the matrix and $\Path_n (G;a,b)$ is the set of
paths of length $n$ from $a$ to $b$.  $(\Vertex(G), \Path(G), \cdot)$ is a category; $F$ is the functor from this category to the multiplicative semigroup of the real numbers which corresponds to the matrix $M$.

@@

\section{Stability criteria}
We call a matrix $M$ stable if all its eigenvalues have negative real
part.  Define the polynomial $p(x) = \det(xI + M)$.  Then, since the
roots of $p$ are minus the eigenvalues of $M$, asking that $M$ be
stable is equivalent to asking that the roots of $p$ all have real
part.  We will formulate conditions for this to happen in terms of the coefficients of $p$.

Let us start by fixing some notation.  Suppose that $M$ is of size $n
\times n$.  Denote the coefficients of $p$ as $s_j$ and its roots as
$r_j$, writing $p(x) = x^n + \sum_{k=1}^n s_k x^{n-k} = \prod_{k=1}^n
(x-r_k)$.

@@

Suppose that we resample $b$.  We want to compute the probability that a random stable matrix will remain stable upon resampling $b$, which is
% \begin{strip}
\begin{align}\label{eq:condprob}
P\left(\begin{pmatrix}
a & b' \\
d & c
\end{pmatrix} \textrm{stable } \bigg| \begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable } \right)
& = \frac{P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable and } \begin{pmatrix}
a & b' \\
d & c
\end{pmatrix} \textrm{stable } \right)}{P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable } \right)}.
\end{align}
The denominator of \ref{eq:condprob} is given by
\begin{align*}
P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable } \right) = \frac{\int_{{ac>bd \atop a+c<0} \atop H^4} da\,db\,dc\,dd\,1}{\int_{H^4} da\,db\,dc\,dd\,1}.
\end{align*}
Since the trace does not involve $b$, the $T<0$ condition will be satisfied automatically and we only need to examine the determinant. Thus, we have the inequalities $ac > b'd$ and $-1 < b' < 1$ in addition to the previous constraints leading to an expression for the numerator of \ref{eq:condprob}
\begin{align*}
P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable and } \begin{pmatrix}
a & b' \\
d & c
\end{pmatrix} \textrm{stable } \right) = \frac{\int_{{{ac>b'd \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,db'\,1}{\int_{H^5} da\,db\,dc\,dd\,db'\,1}.
\end{align*}
The analogous equation for resampling $a$ is
\begin{align*}
P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable and } \begin{pmatrix}
a' & b \\
d & c
\end{pmatrix} \textrm{stable } \right) = \frac{\int_{{{{a'c>bd \atop a' + c < 0} \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,da'\,1}{\int_{H^5} da\,db\,dc\,dd\,da'\,1}.
\end{align*}

@@

\input{tablenographs}

@@

\input{tablewithgraphs}

>>>>

%------------------------------------------------------------------------------
% Beginning of paper.tex
%------------------------------------------------------------------------------
%
% AMS-LaTeX version 2 sample file for journals, based on amsart.cls.
%
%        ***     DO NOT USE THIS FILE AS A STARTER.      ***
%        ***  USE THE JOURNAL-SPECIFIC *.TEMPLATE FILE.  ***
%
% Replace amsart by the documentclass for the target journal, e.g., tran-l.
%
\documentclass{amsart}

@@

\section{Stability analysis of biological networks}\label{sec:stabanalbn}

Consider an n-dimensional vector
$$
x(t) = (x_1(t), \ldots x_n(t)) = \vec{x}(t)
$$
whose components are solutions to the arbitrary first order system
$$
\frac{dx_i(t)}{dt} = F_i(\vec{x}(t)), \; (i=1,\ldots,n)
$$
where $F_i$ represent, potentially nonlinear, functions of the given vector of state variables. We can associate a directed graph $G_X$ that describes the manner in which each of the variables depends upon one another given by an adjacency matrix $\adj(G_X)$ where
 \begin{displaymath}
   \adj(G_X)_{ij} = \left\{
     \begin{array}{ll}
       1, & F_i \hbox{ depends on } x_j\\
       0, & F_i \hbox{ does not depend on } x_j
     \end{array}
   \right.
\end{displaymath}
For example, given the system $X = \{\{x_1,x_2\},\{F_1,F_2\}\}$
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}
we have the graph
\begin{center}
\input{tex/examplehamdyngraph}
\end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$

@@

\section{System graph and decomposition into strongly connected components}
We will show how the stability of the dynamical system associated to Jacobian $A$ is characterized by properties of the associated graph $G_A$. The characteristic polynomial $\chi(A)$ associated to $A$ can be factored according to the strongly connected components of the graph $G$. $$\frac{\chi(M)'(z)}{\chi(M)(z)} = \tr (zI - M)^{-1}$$ where, $\chi(M)(z) = \det (zI - M)$ is the chararacteristic polynomial of the matrix $M$ with $z$ a complex variable.  This formula is readily verified when $M$ is diagonal and has the nice property of being additive over decompositions of the matrix. Making the change of variables $x=\frac{1}{z}$, we have $$\tr (I - xM)^n = \sum_{n=0}^\infty x^n \sum_{a \in \Vertex(G)} \sum_{p \in \Path_n (G)(a,a)} F(p)$$
where $G$ is the graph of the matrix and $\Path_n (G;a,b)$ is the set of
paths of length $n$ from $a$ to $b$.  $(\Vertex(G), \Path(G), \cdot)$ is a category; $F$ is the functor from this category to the multiplicative semigroup of the real numbers which corresponds to the matrix $M$.

@@

\section{Stability criteria}
Suppose that $M$ is an $n \times n$ matrix with real coefficients. Motivated by the discussion of Section \ref{sec:stabanalbn}, we will call the matrix $M$ stable if all its eigenvalues have negative real part.  Define the characteristic polynomial as $\chi(M)(z) = \det(zI - M)$. Denote the coefficients of $\chi(M)$ as $s_j$ and its roots as $r_j$, writing $p(x) = x^n + \sum_{k=1}^n s_k x^{n-k} = \prod_{k=1}^n (x-r_k)$.  Then, since the roots of $\chi$ are the eigenvalues of $M$, asking that $M$ be stable is equivalent to asking that the roots of $\chi$ all have negative real part.  We will formulate conditions for this to happen in terms of the coefficients of $\chi$.

Assume that $M$ is stable and that it has $2m$ complex roots which come in complex conjugate pairs and $n-m$ real roots.  Let $r_1 \ldots, r_{2m}$ be the complex roots with $r_{2k+1}$ being the complex conjugate of $r_{2k}$ and let $r_{2m+1}, \ldots,  r_{n}$ be the real roots. Then we can write the factorization as follows:
\[
\chi(M)(z) = \prod_{k=1}^{m} (z^2 - (r_{2k} + r_{2k+1}) z + r_{2k} r_{2k+1})
             \prod_{k=2m+1}^{n} (z - r_k)
\]
Since $r_{2k}$ and $r_{2k+1}$ are complex conjugates, $r_{2k} + r_{2k+1} =
\Re r_{2k} = \Re r_{2k+1}$ and $r_{k} r_{2k+1} > 0$.   Since all the
roots have negative real part, all the coefficients of the terms in
the products above are positive so, when we multiply out the product,
we will have all positive terms.  Hence, we conclude that a necessary
condition for $M$ to be stable is that $s_k > 0$ for all $k$.

@@

Suppose that we resample $b$.  We want to compute the probability that a random stable matrix will remain stable upon resampling $b$, which is
% \begin{strip}
\begin{align}\label{eq:condprob}
P\left(\begin{pmatrix}
a & b' \\
d & c
\end{pmatrix} \textrm{stable } \bigg| \begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable } \right)
& = \frac{P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable and } \begin{pmatrix}
a & b' \\
d & c
\end{pmatrix} \textrm{stable } \right)}{P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable } \right)}.
\end{align}
The denominator of \ref{eq:condprob} is given by
\begin{align*}
P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable } \right) = \frac{\int_{\genfrac{}{}{0pt}{}{\genfrac{}{}{0pt}{}{ac>bd}{a+c<0}}{H^4}} da\,db\,dc\,dd\,1}{\int_{H^4} da\,db\,dc\,dd\,1}.
\end{align*}
Since the trace does not involve $b$, the $T<0$ condition will be satisfied automatically and we only need to examine the determinant. Thus, we have the inequalities $ac > b'd$ and $-1 < b' < 1$ in addition to the previous constraints leading to an expression for the numerator of \ref{eq:condprob}
\begin{align*}
P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable and } \begin{pmatrix}
a & b' \\
d & c
\end{pmatrix} \textrm{stable } \right) = \frac{\int_{{{ac>b'd \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,db'\,1}{\int_{H^5} da\,db\,dc\,dd\,db'\,1}.
\end{align*}
The analogous equation for resampling $a$ is
\begin{align*}
P\left(\begin{pmatrix}
a & b \\
d & c
\end{pmatrix} \textrm{stable and } \begin{pmatrix}
a' & b \\
d & c
\end{pmatrix} \textrm{stable } \right) = \frac{\int_{{{{a'c>bd \atop a' + c < 0} \atop ac>bd} \atop a+c<0} \atop H^5} da\,db\,dc\,dd\,da'\,1}{\int_{H^5} da\,db\,dc\,dd\,da'\,1}.
\end{align*}

@@

\input{tex/tablenographs}

@@

\input{tex/tablewithgraphs}
